Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2023-07-20 src/docs_paper/paper.Rmd
Improve writing of AGHQ section #55

Quadrature rules can be used to approximate integrals using a set of nodes $\z \in \mathcal{Q}$ and a weighting function $\omega: \mathcal{Q} \to \mathbb{R}$.
Gauss-Hermite quadrature [GHQ; @davis1975methods] is a quadrature rule, where in the univariate case the nodes are selected as zeros of the $k$th Hermite polynomial
\begin{equation}
\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}.
\end{equation}
The corresponding weights $\omega: \mathcal{Q}(1, k) \to \mathbb{R}$ are given by $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$, where $\phi(\cdot)$ is a standard Gaussian density.
GHQ is attractive because it is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$.
We expect posterior distributions to be relatively well described by this class of functions.
To integrate over $\btheta$ we require a multivariate GHQ rule, which is typically obtained using the product rule such that $\z = (z_1, \ldots, z_m) \in \mathcal{Q}(m, k) = \mathcal{Q}(1, k)^m$ and $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$.
In adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] the nodes are shifted and rotated to suit the particular integrand.
Repositioning the nodes is especially important in statistical quadrature problems, where the integral depends on data $\y$ and regions of high density are not known in advance.

Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature, then an AGHQ estimate of the normalising constant $p(\y)$ based on the Laplace approximation is
\begin{equation}
p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z). \label{eq:aghq}
\end{equation}
The unadapted nodes are shifted by the mode and rotated by a decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
Two alternatives [@jackel2005note] are
1. the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
2. the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
Equation \ref{eq:aghq} may then be used to normalise the Laplace approximation
\begin{equation}
\tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
\end{equation}
To obtain inferences for the latent field (Equation \ref{eq:inla1}) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
\begin{equation}
\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z). \label{eq:nest}
\end{equation}
Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.

@@

Use of the product rule grid described above requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points.
This quickly becomes intractable as $m$ increases for $k > 1$.
An alternative is to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$.
We may then define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ to be a GHQ grid with possible variable levels of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
For example, for $m = 2$ and $s = 1$ then $\bk = (k, 1)$.
In combination with use of the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ.
We refer to this approach as PCA-AGHQ, with corresponding estimate of the normalising constant given by
\begin{equation}
\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}} \hat{\mathbf{\Lambda}}_{\texttt{LA}}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z),
\end{equation}
where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\omega(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
Panel C of Figure \ref{fig:aghq} illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.

>>>>

Let $\z \in \mathcal{Q}(m, k)$ be $m$-dimensional Gauss-Hermite quadrature [GHQ; @davis1975methods] rule with $k$ nodes per dimension constructed using the product rule such that $\mathcal{Q}(m, k) = \mathcal{Q}(1, k) \times \cdots \times \mathcal{Q}(1, k)$ where
\begin{equation}
\mathcal{Q}(1, k) = \{z: H_k(z) = (-1)^k \exp(z^2 / 2) \frac{\text{d}}{\text{d}z^k} \exp(-z^2 / 2) = 0\}, \\
\end{equation}
with $\phi(\cdot)$ is a standard Gaussian density.
The corresponding weighting function $\omega: \mathcal{Q}(m, k) \to \mathbb{R}$ is given by $\omega(\z) = \prod_{j = 1}^m \omega(z_j)$ where $\omega(z) = k! / [H_{k + 1}(z)]^2 \phi(z)$.
For $k = 1$ GHQ corresponds to a Laplace approximation.
Further GHQ is exact for functions which are a Gaussian density multiplied by a polynomial of total order no more than $2k - 1$, a class of functions we expect the posterior to be close to.

Let $\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA}) = - \partial^2 \log p_\texttt{LA}(\hat{\btheta}_\texttt{LA}, \y)$ be the curvature at the mode $\hat{\btheta}_\texttt{LA}$ and $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1} = \hat{\mathbf{P}}_\texttt{LA} {\hat{\mathbf{P}}_\texttt{LA}}^\top$ be a matrix decomposition of the inverse curvature.
An adaptive Gauss-Hermite quadrature [AHGQ; @naylor1982applications; @tierney1986accurate] estimate of the normalising constant $p(\y)$ based on the Laplace approximation is given by
\begin{equation}
p(\y) \approx \int_{\btheta} \tilde p_\texttt{LA}(\btheta, \y) \approx \tilde p_\texttt{AGHQ}(\y) = |\hat{\mathbf{P}}_\texttt{LA}|\sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z). \label{eq:aghq}
\end{equation}
The unadapted nodes are shifted by the mode and rotated by a matrix decomposition of the inverse curvature such that $\z \mapsto \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}$.
Repositioning the nodes is crucial for statistical quadrature problems like ours, where the integral depends on data $\y$ and regions of high density are not known in advance.
Two alternatives for the matrix decomposition [@jackel2005note] are
1. the Cholesky decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{L}}_\texttt{LA}$, where $\hat{\mathbf{L}}_\texttt{LA}$ is lower triangular, and
2. the spectral decomposition $\hat{\mathbf{P}}_\texttt{LA} = \hat{\mathbf{E}}_\texttt{LA} \hat{\mathbf{\Lambda}}_\texttt{LA}^{1/2}$, where $\hat{\mathbf{E}}_\texttt{LA} = (\hat{\mathbf{e}}_{\texttt{LA}, 1}, \ldots \hat{\mathbf{e}}_{\texttt{LA}, m})$ contains the eigenvectors of $[\hat{\Hb}_\texttt{LA}(\hat{\btheta}_\texttt{LA})]^{-1}$ and $\hat{\mathbf{\Lambda}}_\texttt{LA}$ is a diagonal matrix containing its eigenvalues $(\hat \lambda_{\texttt{LA}, 1}, \ldots, \hat \lambda_{\texttt{LA}, m})$.
This estimate may be used to normalise the Laplace approximation
\begin{equation}
\tilde p_\texttt{LA}(\btheta \, | \, \y) = \frac{\tilde p_\texttt{LA}(\btheta, \y)}{\tilde p_\texttt{AGHQ}(\y)}.
\end{equation}
To obtain inferences for the latent field (Equation \ref{eq:inla1}) we reuse the adapted nodes and weights [@rue2009approximate; @stringer2022fast]
\begin{equation}
\tilde p(\x \, | \, \y) = |\hat{\mathbf{P}}_\texttt{LA}| \sum_{\z \in \mathcal{Q}(m, k)} \tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y) \tilde p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z). \label{eq:nest}
\end{equation}
Samples from this mixture of Gaussians may be obtained by drawing a node $\z$ with multinomial probabilities $\lambda(\z) = |\hat{\mathbf{P}}_\texttt{LA}| p_\texttt{LA}(\hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA} \, | \, \y) \omega(\z)$, then drawing from the corresponding Gaussian $\tilde p_\texttt{G}(\x \, | \, \hat{\mathbf{P}}_\texttt{LA} \z + \hat{\btheta}_\texttt{LA}, \y)$.

@@

Use of the product rule grid described above requires $|\mathcal{Q}(m, k)| = k^m$ quadrature points.
This quickly becomes intractable as $m$ increases for $k > 1$.
An alternative is to let $\bk = (k_1, \ldots, k_m)$ be a vector of levels for each dimension of $\btheta$.
We may then define $\mathcal{Q}(m, \bk) = \mathcal{Q}(1, k_1) \times \cdots \times \mathcal{Q}(1, k_m)$ to be a GHQ grid with possible variable levels of size $|\mathcal{Q}(m, \bk)| = \prod_{j = 1}^m k_j$.
Let $\mathcal{Q}(m, s, k)$ correspond to $\mathcal{Q}(m, \bk)$ with choice of levels $k_j = k, j \leq s$ and $k_j = 1, j > s$ for some $s \leq m$.
For example, for $m = 2$ and $s = 1$ then $\bk = (k, 1)$.
In combination with use of the spectral decomposition, this choice of levels is analogous to a principal components analysis (PCA) approach to AGHQ.
We refer to this approach as PCA-AGHQ, with corresponding estimate of the normalising constant given by
\begin{equation}
\tilde p_\texttt{PCA}(\y) = |\hat{\mathbf{E}}_{\texttt{LA}} \hat{\mathbf{\Lambda}}_{\texttt{LA}}^{1/2}|\sum_{\z \in \mathcal{Q}(m, s, k)} \tilde p_\texttt{LA}(\hat{\mathbf{E}}_{\texttt{LA}, s} \hat{\mathbf{\Lambda}}_{\texttt{LA}, s}^{1/2} \z + \hat{\btheta}_\texttt{LA}, \y) \omega(\z),
\end{equation}
where $\hat{\mathbf{E}}_{\texttt{LA}, s}$ is an $m \times s$ matrix containing the first $s$ eigenvectors, $\hat{\mathbf{\Lambda}}_{\texttt{LA}, s}$ is the $s \times s$ diagonal matrix containing the first $s$ eigenvalues, and $\omega(\z) = \prod_{j = 1}^s \omega_s(z_j) \times \prod_{j = s + 1}^d \omega_1(z_j)$.
Panel C of Figure \ref{fig:aghq} illustrates PCA-AGHQ for a case when $m = 2$ and $s = 1$.
As AGHQ with $k = 1$ corresponds to the Laplace approximation, PCA-AGHQ can be interpreted as performing AGHQ on the first $s$ principal components of the inverse curvature, and a Laplace approximation on the remaining $m - s$ principal components.
Inference for the latent field follows analogously to Equation \ref{eq:nest}.
